---
title: "Supplement to Hassell, Lawton and May 1976"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

\subsection{Eqn. 1}

Hassell, Lawton \& May (1976) base their work on the model
\begin{equation}
	N_{t+1}=\frac{\lambda N_t}{(1+\alpha N_t)^{-\beta}}
\end{equation}
which might be easier to visualize as
\begin{equation*}
		\frac{N_{t+1}}{N_t}=\frac{\lambda }{(1+\alpha N_t)^{-\beta}}.
\end{equation*}

Observe that when $a=0$ or $\beta=0$ this reduces to density-independent growth
\begin{equation*}
	\frac{N_{t+1}}{N_t}=\lambda.
\end{equation*}

When $a>0$ or $\beta>0$ (as assumed in the paper), we can see in the figures below that the greater their values (i.e. the greater that $a$ or $\beta$ are), the more curved the function becomes.  More specifically, $a$ determines where the inflection point of the curve is (larger values create an inflection at lower values of $N_t$), and $\beta$ determines the strength of the density-dependence (the steepness (slope) of the curve).

```{r fig1, fig.height = 3, fig.width = 8, echo=FALSE}
foo<-function(x){lambda/((1+a*x)^b)}
par(mfrow=c(1,3))
lambda=2;a=0;b=0;
curve(foo,0,10,ylab=expression(N[t+1]/N[t]),xlab=expression(N[t]),main=bquote(paste(lambda==.(lambda),', ',a==.(a),', ',beta==.(b))))

lambda=2;a=0.1;b=1;
curve(foo,0,10,ylab=expression(N[t+1]/N[t]),xlab=expression(N[t]),main=bquote(paste(lambda==.(lambda),', ',a==.(a),', ',beta==.(b))))

lambda=2;a=2;b=1;
curve(foo,0,10,ylab=expression(N[t+1]/N[t]),xlab=expression(N[t]),main=bquote(paste(lambda==.(lambda),', ',a==.(a),', ',beta==.(b))))
```

\pagebreak
\subsection*{Eqn. 2}
They also give
\begin{equation}
\log \frac{N_t}{N_s}=\beta \log(1+a N_t)
\end{equation}
where $N_t$ is the initial number of individuals at time $t$, and $N_s$ is the number of survivors at time $t+1$ (\emph{not} the total number of individuals, $N_{t+1}$).

So how is eqn. 2 related to eqn. 1?  Answer: ``simply'', according to HLM in the line after eqn. 2!

First, let's note that
\begin{equation*}
x^{-1} = \frac{1}{x}
\end{equation*}
and that
\begin{equation*}
	\log(x^a) = a \log(x).
\end{equation*}
So now,
\begin{align}
	N_{t+1}&=\lambda N_t (1+a N_t)^{-\beta} \\
		& = \lambda N_s\\
	\implies \frac{N_s}{N_t}&=(1+a N_t)^{-\beta} \\
	\implies  \frac{N_t}{N_s}&=(1+a N_t)^{\beta} \\
	\implies \log \frac{N_s}{N_t}&=\beta \log(1+a N_t)
\end{align}

